---
title: "ATT_under_positivity_violations"
author: "Skyler Shapiro"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("/Users/skylershapiro/duke/DCRI/OWATT_simulations_skyler/simulation_results/bootvar/OWATT_functions.R")
library(knitr)
```

### Define three overlap scenarios.
```{r}
# Define three levels of positivity
good_kappas <- c(0,0.08,0.2,0.1,0.13)
adequate_kappas <- c(0,.4,0.8,0.4,0.35)
# poor_kappas <- c(0,1,1.9,.87,0.8)
poor_kappas <- c(0,0.1,0.2,0.1,.1)

```

#### EDA: good overlap
```{r cars}
# Generate data
N = 2000

X <- make_data(N)

#one setting: ks <- c(0,0.4,0.2,0.1,.1) als <- c(3,.95,.25, 2.5, .3)
# another setting <- alphas = c(4,1.95,0.25,0.1, 1.8), kappas = c(0,0.1,0.2,0.1,.1)

z_good <- make_treatment(X, N,kappas = good_kappas)
z_adequate <- make_treatment(X, N, alphas = c(3,1.95,0.25,0.1, 1.8),kappas = adequate_kappas)
z_poor <- make_treatment(X, N, alphas = c(4,1.95,0.25,0.1, 1.8), kappas = c(0,0.1,0.2,0.1,.1))

Y_good <- make_outcome(X, z_good, N)
Y_adequate <- make_outcome(X, z_adequate, N)
Y_poor <- make_outcome(X, z_poor, N)
```


```{r}
# structural positivity violation
ks <- c(0,0.4,0.1,0.4,.1)
als <- c(3,.95,.25, 2.5, .3)
plot_overlap(kappas = ks, alphas = als)
tab <- table(z_poor <- make_treatment(X, N, alphas = als, kappas = ks))
kable(cbind(Trt = 1:5, Freq = tab, Percent = round(tab / N * 100, 2)))
```

```{r}
# no structural positivity violation
ks <- c(0,.2,.3, .2,.3)
als <- c(2.2,.95,.25, 1, .3)
plot_overlap(kappas = ks, alphas = als)
tab <- table(z_poor <- make_treatment(X, N, alphas = als, kappas = ks))
kable(cbind(Trt = 1:5, Freq = tab, Percent = round(tab / N * 100, 2)))
```


```{r}
## VALIDATE PROPENSITY PLOTS WITH HISTOGRAM/COUNTS TO VERIFY STRUCTURAL POSITIVITY VIOLATION
```


```{r}
plot_overlap(good_kappas)
```

```{r}
tab <- table(make_treatment(X, N, alphas = c(-2, 4.0, -4.0, 0.5, 0.5),kappas = c(0, 0.1,0.2, .2, .2, .2)))
kable(cbind(Trt = 1:5, Freq = tab, Percent = round(tab / N * 100, 2)))
```

#### EDA: adequate overlap
```{r}
plot_overlap(kappas = c(0,0.1,0.2,0.1,.1), alphas = c(4,1.95,0.25,0.1, 1.8))
```

```{r}
tab <- table(z_poor)
kable(cbind(Trt = 1:5, Freq = tab, Percent = round(tab / N * 100, 2)))
```

#### EDA: poor overlap
```{r}
plot_overlap(poor_kappas, alphas = c(4,1.95,0.25,0.1, 1.8))
```

```{r}
tab <- table(z_poor)
kable(cbind(Trt = 1:5, Freq = tab, Percent = round(tab / N * 100, 2)))
```

### Define weighting procedures:

1. **ipw:** selection function: $h = 1$
2. **trimming:** selection function: $h = I_{[\alpha, (1-\alpha)]}(e_j)$. Generalized propensity scores are re-estimated after trimming to improve finite sample performance per Li and Li, 2019.
3. **truncation**  $h = \begin{cases}  e_j & e_j \in [\alpha, 1-\alpha]\\ \alpha & e_j < \alpha \\ 1-\alpha & e_j > 1-\alpha  \end{cases}$
4. **influence trimming** $h = I_{(-\infty, \alpha]}(w_i)$ with $w_i$ being the vector of ipw weights for patients in group $i$.
5. **generalized matching weights** $h_i = \min \{ e_{i1}, e_{i2}, \dots, e_{iJ} \}$
6. **generalized overlap weights** $h  = \left( \sum_{k=1}^J \frac{1}{e_k(X)} \right)^{-1}$
7. **Shannon's entropy weight**  $h= -[e_j \log(e_j) + e_i \log(e_i)]$
8. **beta weights**  $h = (e_j \cdot e_i)^{\nu-1}, \quad \nu \ge 2$

```{r}
## ADD PLOT of KERNELS
```


### Examine weight distributions of IPW and OW under three overlap scenarios
```{r}
z_factor <- factor(z_good, levels = 1:5)
fit <- nnet::multinom(z_factor ~ X, trace = FALSE)
e_full <- predict(fit, type = "probs")   # N x 5 matrix, columns = groups 1-5
e_j <- e_full[, 1]

ipw_mat_good <- sapply(2:5, function(i) e_j / e_full[, i])
ow_mat_good <- sapply(2:5, function(i)( e_j * (1 / rowSums(1 / e_full) ) / e_full[, i]))
colnames(ow_mat_good) <- paste0(2:5, "v1")
colnames(ipw_mat_good) <- paste0(2:5, "v1")

z_factor <- factor(z_adequate, levels = 1:5)
fit <- nnet::multinom(z_factor ~ X, trace = FALSE)
e_full <- predict(fit, type = "probs")   # N x 5 matrix, columns = groups 1-5
e_j <- e_full[, 1]

ipw_mat_adequate<- sapply(2:5, function(i) e_j / e_full[, i])
ow_mat_adequate <- sapply(2:5, function(i)( e_j * (1 / rowSums(1 / e_full) ) / e_full[, i]))
colnames(ipw_mat_adequate) <- paste0(2:5, "v1")
colnames(ow_mat_adequate) <- paste0(2:5, "v1")

z_factor <- factor(z_poor, levels = 1:5)
fit <- nnet::multinom(z_factor ~ X, trace = FALSE)
e_full <- predict(fit, type = "probs")   # N x 5 matrix, columns = groups 1-5
e_j <- e_full[, 1]

ipw_mat_poor<- sapply(2:5, function(i) e_j / e_full[, i])
ow_mat_poor <- sapply(2:5, function(i)( e_j * (1 / rowSums(1 / e_full) ) / e_full[, i]))
colnames(ipw_mat_poor) <- paste0(2:5, "v1")
colnames(ow_mat_poor) <- paste0(2:5, "v1")

```

```{r}
rbind(summary(ipw_mat_good)) |> kable()
```

```{r}
rbind(summary(ipw_mat_adequate)) |> kable()
```

```{r}
rbind(summary(ipw_mat_poor)) |> kable()
```

```{r}
rbind(summary(ow_mat_good)) |> kable()
```

```{r}
rbind(summary(ow_mat_adequate)) |> kable()
```

```{r}
rbind(summary(ow_mat_poor)) |> kable()
```

```{r}
PSDfun = function(Z, D, covM, h, w) {
  # Z: treatment value
  # D: matrix of treatment indicators
  # covM: covariate matrix, n by p
  # h: value of the h function, n by 1
  # w: matrix of weights, n by J

  p = ncol(covM)
  J = ncol(D)
  psdM = matrix(NA, J, p)
  
  for (k in 1:p) {
    x = as.numeric(covM[, k])
    xbar.p = sum(rowSums(D * x * h)) / sum(rowSums(D * h))
    xbar.j = colSums(D * x * w) / colSums(D * w)
    s2x = mean(tapply(x, Z, var))
    psdM[, k] = abs(xbar.j - xbar.p) / sqrt(s2x)
  }
  return(t(psdM))
}

# create contrast matrix
J = 3
Jc = choose(J, 2) # number of contrasts
loc = t(combn(J, 2))
C = matrix(0, Jc, J)

for(i in 1:Jc) {
  C[i, loc[i, 1]] = 1
  C[i, loc[i, 2]] = -1
}

# Absolute standardized difference
ASDfun = function(Z, D, C, covM, h, w) {
  # Z: treatment value
  # D: matrix of treatment indicators
  # C: contrast matrix
  # covM: covariate matrix, n by p
  # h: value of the h function
  # w: matrix of weights, n by J

  p = ncol(covM)
  Jc = nrow(C)
  asdM = matrix(NA, Jc, p)

  # loop through each pair
  for (k in 1:p) {
    x = as.numeric(covM[, k])
    xbar.j = colSums(D * x * w) / colSums(D * w)
    s2x = mean(tapply(x, Z, var))
    asdM[, k] = abs(C %*% xbar.j) / sqrt(s2x)
  }
  return(t(asdM))
}
```


```{r}
# Setup dimensions
n <- nrow(X)
J <- 5 # Number of groups
D <- model.matrix(~factor(z_poor) - 1) # N x 5 indicator matrix
h <- rep(1, n)                        # Target: Overall population (ATE)
w <- cbind(1,ow_mat_poor)                    # Your calculated overlap weights

unweighted_results <- round(PSDfun(Z = z_poor, D = D, covM = X, h = h, w = rep (1 , N)), 3)
print(unweighted_results)

ipw_results <- round(PSDfun(Z = z_poor, D = D, covM = X, h = h, w = cbind(1, ipw_mat_poor)), 3)
print(unweighted_results)

psd_results <- round(PSDfun(Z = z_poor, D = D, covM = X, h = h, w = w), 3)
print(psd_results)
```


```{r fig.width=18, fig.height=10}
library(ggplot2)
library(tidyr)
library(patchwork)
library(scales)

group_colors <- c("#E69F00", "#56B4E9", "#CC79A7", "#D55E00", "#009E73")
ipw_long_good <- pivot_longer(as.data.frame(ipw_mat_good), everything())
ipw_long_adequate <- pivot_longer(as.data.frame(ipw_mat_adequate), everything())
ipw_long_poor <- pivot_longer(as.data.frame(ipw_mat_poor), everything())

p1 <- ggplot(ipw_long_good, aes(name, value, fill = name)) +
  geom_boxplot() + 
  scale_fill_manual(values = group_colors)

ipw_long_adequate$value <- pmax(ipw_long_adequate$value, 1e-6)
p2 <- ggplot(ipw_long_adequate, aes(name, value, fill = name)) +
    geom_boxplot() +
    scale_fill_manual(values = group_colors) +
    scale_y_log10(
        breaks = trans_breaks("log10", function(x) 10^x),
        labels = label_number()) # Or use label_log() for 10^x formatting 
        
        
ipw_long_poor$value <- pmax(ipw_long_poor$value, 1e-6)
p3 <- ggplot(ipw_long_poor, aes(name, value, fill = name)) +
    geom_boxplot() +
    scale_fill_manual(values = group_colors) +
    scale_y_log10(
        breaks = trans_breaks("log10", function(x) 10^x),
        labels = label_number() # Or use label_log() for 10^x formatting
    )

p1|p2|p3
```

```{r fig.width=18, fig.height=10}
group_colors <- c("#E69F00", "#56B4E9", "#CC79A7", "#D55E00", "#009E73")
ow_long_good <- pivot_longer(as.data.frame(ow_mat_good), everything())
ow_long_adequate <- pivot_longer(as.data.frame(ow_mat_adequate), everything())
ow_long_poor <- pivot_longer(as.data.frame(ow_mat_poor), everything())

p1 <- ggplot(ow_long_good, aes(name, value, fill = name)) +
  geom_boxplot() + 
  scale_fill_manual(values = group_colors)

ow_long_adequate$value <- pmax(ow_long_adequate$value, 1e-6)
p2 <- ggplot(ow_long_adequate, aes(name, value, fill = name)) +
    geom_boxplot() +
    scale_fill_manual(values = group_colors) #+
    # scale_y_log10(
    #     breaks = trans_breaks("log10", function(x) 10^x),
    #     labels = label_number()) # Or use label_log() for 10^x formatting 
        
        
ow_long_poor$value <- pmax(ow_long_poor$value, 1e-6)
p3 <-ggplot(ow_long_poor, aes(name, value, fill = name)) +
    geom_boxplot() +
    scale_fill_manual(values = group_colors) #+
    # scale_y_log10(
    #     breaks = trans_breaks("log10", function(x) 10^x),
    #     labels = label_number() # Or use label_log() for 10^x formatting
    # )

p1|p2|p3
```

```{r}

```


```{r}

```

